Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-21 paper.Rnw
Adding basic WA-volume plot

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
% 
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings explored
are: atlas library size (1-9), template library size (1-20), registration method
(\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  In each validation round, both a
\mb{} and \multiatlas segmentation is produced.  A total of $10 \times 60 \times
9 \times 20 \times 2 \times 3 = \Sexpr{10*60*9*20*2*3}$ validation rounds
were conducted and resulting segmentations analysed.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
<<WAval-prep, cache=T>>=
data = subset(read.csv('data/WAval.csv'))
casted = dcast(melt(data, id.vars=c("file","version")), file + variable ~ version)
casted = rename(casted, c("variable"="region"))

error = ddply(casted, c("file", "region"), function(df) {
  with(df, 
  data.frame(
    resampled = (gold_0.9mm - gold_0.3mm) / gold_0.3mm * 100, 
    mb_bravo  = (mb_bravo   - gold_0.3mm) / gold_0.3mm * 100, 
    mb_0.9mm  = (mb_0.9mm   - gold_0.3mm) / gold_0.3mm * 100))})

levels(error$region) <-list("CA1"="X1", "CA1"="X101", 
                            "Subiculum"="X2","Subiculum"="X102", 
                            "CA4/DG"="X4","CA4/DG"="X104", 
                            "CA2/CA3"="X5","CA2/CA3"="X105", 
                            "SR/SL/SM"="X6","SR/SL/SM"="X106")
melted = melt(error,id.vars=c("file", "region"))
melted = rename(melted, c("variable"="measure", "value"="percenterr"))
levels(melted$measure) <- list("Subsampled atlas labels"="resampled",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

error.t.tests = ddply(error, c("region"), function (df) { 
  data.frame( 
    bravo = round(t.test(df$resampled, df$mb_bravo, paired=T)$p.value,3), 
    atlas = round(t.test(df$resampled, df$mb_0.9mm, paired=T)$p.value,3)
  )}) 
@
This experiment explores \mb{} segmentation of hippocampal subfields. To achieve
this, a leave-one-out validation was conducted in which lower-resolution images
($0.9mm^3$ voxels) of each Winterburn atlas subject were segmented using the
remaining high-resolution Winterburn atlas subjects' images.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<WAval-vol-boxplot,cache=T,fig.width=6,fig.height=4>>=
ggplot(melted, aes(y=percenterr,x=measure,colour=measure)) +
  facet_grid(.~region) + 
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() + 
  geom_point(group=1, position="dodge") + 
  labs(y="Unilateral subregion volume percent error", x="Subfield") +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  scale_colour_discrete("") + 
  theme(legend.direction = "vertical", legend.position = "bottom")
@
  \caption{{\bf Percent error in unilateral subfield volume.} 
  Percent error is measured against the volumes of the unmodified Winterburn
  atlas subfield segmentations.
  {\bf Subsampled atlas labels} volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$ voxels.  
  {\bf MAGeT on subsampled atlas images} volumes are \mb{} segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT on T1 BRAVO} volumes are \mb{} segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 1yr Complete Dataset Package Totals & Failures
%                        
<<ADNI-xvals-pvalue, echo=F, results="asis",cache=FALSE>>=
latex(error.t.tests, file="", size="scriptsize",landscape=FALSE, ctable=TRUE,
    caption="\\textbf{t-test p-values whem comparing to resampled percent error.}", 
    caption.loc = c("top"), rowname=NULL, 
    label="tab:resampled-percent-error")
@

@@

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup",fig.width=5,fig.height=5>>=
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
mean_sz_vols = ddply(SZ_volumes, c('Subject', 'Method'), function (df) {
  data.frame(Volume = mean(df$Volume))})

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
% 
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

@@

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings explored
are: atlas library size (1-9), template library size (1-20), registration method
(\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  In each validation round, both a
\mb{} and \multiatlas segmentation is produced.  A total of $10 \times 60 \times
9 \times 20 \times 2 \times 3 =$ \Sexpr{10*60*9*20*2*3} validation rounds
were conducted and resulting segmentations analysed.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
<<WAval-prep, cache=T>>=
wa.xval.data = subset(read.csv('data/WAval.csv'))
wa.xval.casted = dcast(melt(wa.xval.data, id.vars=c("file","version")), file + variable ~ version)
wa.xval.casted = rename(wa.xval.casted, c("variable"="region"))
levels(wa.xval.casted$region) <-list("CA1"="X1", "CA1"="X101", 
                            "Subiculum"="X2","Subiculum"="X102", 
                            "CA4/DG"="X4","CA4/DG"="X104", 
                            "CA2/CA3"="X5","CA2/CA3"="X105", 
                            "SR/SL/SM"="X6","SR/SL/SM"="X106")
@
This experiment explores \mb{} segmentation of hippocampal subfields. To achieve
this, a leave-one-out validation was conducted in which lower-resolution images
($0.9mm^3$ voxels) of each Winterburn atlas subject were segmented using the
remaining high-resolution Winterburn atlas subjects' images.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<WAval-vol-boxplot,cache=T,fig.width=6,fig.height=4>>=
wa.xval.error = ddply(wa.xval.casted, c("file", "region"), function(df) {
  with(df, 
  data.frame(
    resampled = (gold_0.9mm - gold_0.3mm) / gold_0.3mm * 100, 
    mb_bravo  = (mb_bravo   - gold_0.3mm) / gold_0.3mm * 100, 
    mb_0.9mm  = (mb_0.9mm   - gold_0.3mm) / gold_0.3mm * 100))})

wa.xval.error.melted = melt(wa.xval.error,id.vars=c("file", "region"))
wa.xval.error.melted = rename(wa.xval.error.melted, c("variable"="measure", "value"="percenterr"))
levels(wa.xval.error.melted$measure) <- list("Subsampled atlas labels"="resampled",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

ggplot(wa.xval.error.melted, aes(y=percenterr,x=measure,colour=measure)) +
  facet_grid(.~region) + 
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() + 
  geom_point(group=1, position="dodge") + 
  labs(y="Unilateral subregion volume percent error", x="Subfield") +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  scale_colour_discrete("") + 
  theme(legend.direction = "vertical", legend.position = "bottom")
@
  \caption{{\bf Percent error in unilateral subfield volume.} 
  Percent error is measured against the volumes of the unmodified Winterburn
  atlas subfield segmentations.
  {\bf Subsampled atlas labels} volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$ voxels.  
  {\bf MAGeT on subsampled atlas images} volumes are \mb{} segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT on T1 BRAVO} volumes are \mb{} segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: volume plots
\begin{figure}[h!]
<<WAval-vol-plots,cache=T,fig.width=6,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.vols.lm = ddply(wa.xval.volumes, c("region", "measure"), function(df) {
   model = lm(volume ~ gold_0.3mm, df)
   pearson = cor(df$volume, df$gold_0.3mm, use='pairwise.complete.obs')
   data.frame(
     eqn  = lm_eqn(model), 
     corr = pearson)})

ggplot(wa.xval.volumes, aes(x=gold_0.3mm, y=volume, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_point() + 
  geom_smooth(method='lm')  + 
  coord_equal() + 
  ylab(expression("Automated Unilateral Volume" (mm^3))) +
  xlab(expression("Manual Unilateral Volume" (mm^3))) +
  geom_text(aes(x=Inf-200, y=Inf-200, label=eqn), parse=TRUE, data=wa.xval.vols.lm, position="dodge") +
  theme(legend.direction = "horizontal", legend.position = "bottom")

@
  \label{fig:WAval-volumes}
\end{figure}

@@

@@

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup",fig.width=5,fig.height=5>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
mean_sz_vols = ddply(SZ_volumes, c('Subject', 'Method'), function (df) {
  data.frame(Volume = mean(df$Volume))})
